---
title: "REDCap Data Extraction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{REDCap Data Extraction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# REDCap Data Extraction with Sardine

This vignette demonstrates how to extract data from REDCap projects using the sardine package's object-oriented approach.

## Setup

First, load the required packages:

```{r setup}
library(sardine)
library(dplyr)
library(tibble)
```

## Environment Configuration

The sardine package uses environment variables for secure credential management:

```{r env_setup, eval=FALSE}
# Create environment template (one-time setup)
create_env_template()

# Edit the .env file with your REDCap credentials:
# REDCAP_URL=https://redcap.example.edu/api/
# REDCAP_TOKEN=YOUR_API_TOKEN_HERE

# Load environment variables
load_env()
```

## Creating a REDCap Project

The core of sardine's approach is the REDCap project object:

```{r project_creation, eval=FALSE}
# Create project from environment variables
project <- redcap_project_from_env()

# Or create directly with credentials
project <- redcap_project(
  url = "https://redcap.example.edu/api/",
  token = "YOUR_API_TOKEN"
)
```

When you create a project:
- Connection is tested automatically
- Full dataset is cached for performance
- Project metadata is retrieved
- Any connection issues cause immediate failure with clear error messages

## Viewing Project Information

```{r project_info, eval=FALSE}
# View project summary
project$info()
# REDCap Project
# ==============
# Title: My Research Study
# URL: https://redcap.example.edu
# Created: 2025-10-03 10:30:00
# Cached Data: 150 records, 25 fields
```

## Accessing All Data

The simplest way to get all your data:

```{r all_data, eval=FALSE}
# Access cached full dataset
all_data <- project$data

# View data structure
glimpse(all_data)

# Basic summary
cat("Records:", nrow(all_data), "\n")
cat("Fields:", ncol(all_data), "\n")
cat("Record ID field:", names(all_data)[1], "\n")
```

## Extracting Specific Fields

For targeted data extraction:

```{r specific_fields, eval=FALSE}
# Extract demographic fields
demographics <- export_records(
  project,
  fields = c("record_id", "age", "gender", "race", "ethnicity")
)

# Extract baseline measurements
baseline <- export_records(
  project,
  fields = c("record_id", "height", "weight", "bp_systolic", "bp_diastolic")
)

# View results
head(demographics)
```

## Extracting Specific Records

To get data for particular participants:

```{r specific_records, eval=FALSE}
# Extract data for specific participants
pilot_participants <- export_records(
  project,
  records = c("001", "002", "003", "004", "005")
)

# Extract records meeting criteria (using cached data)
high_risk <- project$data %>%
  filter(risk_score > 75) %>%
  pull(record_id)

high_risk_data <- export_records(
  project,
  records = high_risk
)
```

## Extracting by Form/Instrument

To get data from specific forms:

```{r by_form, eval=FALSE}
# Get all instruments/forms
instruments <- export_instruments(project)
print(instruments)

# Extract data from specific form
baseline_survey <- export_records(
  project,
  forms = "baseline_demographics"
)

# Extract from multiple forms
core_data <- export_records(
  project,
  forms = c("demographics", "medical_history", "current_medications")
)
```

## Working with Longitudinal Projects

For projects with multiple events:

```{r longitudinal, eval=FALSE}
# Check if project is longitudinal
events <- export_events(project)

if (nrow(events) > 0) {
  cat("Longitudinal project with", nrow(events), "events\n")
  
  # Extract data from specific event
  baseline_event <- export_records(
    project,
    events = "baseline_arm_1"
  )
  
  # Extract from multiple events
  followup_data <- export_records(
    project,
    events = c("month_3_arm_1", "month_6_arm_1", "month_12_arm_1")
  )
  
  # Combine specific fields and events
  vital_signs_over_time <- export_records(
    project,
    fields = c("record_id", "height", "weight", "bp_systolic", "bp_diastolic"),
    events = c("baseline_arm_1", "month_3_arm_1", "month_6_arm_1")
  )
}
```

## Data Refresh

When you need fresh data from REDCap:

```{r refresh, eval=FALSE}
# Refresh cached data
project$refresh()

# The cached data is now updated
fresh_data <- project$data
```

## Best Practices

### 1. Use Cached Data When Possible
```{r best_practices_cache, eval=FALSE}
# Good: Use cached data for exploration
all_data <- project$data
summary_stats <- all_data %>%
  summarise(
    n_participants = n_distinct(record_id),
    mean_age = mean(age, na.rm = TRUE),
    complete_records = sum(complete.cases(.))
  )
```

### 2. Use Targeted Extraction for Specific Analyses
```{r best_practices_targeted, eval=FALSE}
# Good: Extract only needed fields for analysis
analysis_data <- export_records(
  project,
  fields = c("record_id", "treatment_group", "outcome_measure", "baseline_score")
)
```

### 3. Handle Missing Data Appropriately
```{r best_practices_missing, eval=FALSE}
# Check for missing data patterns
missing_summary <- project$data %>%
  summarise_all(~sum(is.na(.))) %>%
  gather(field, missing_count) %>%
  arrange(desc(missing_count))

print(missing_summary)
```

### 4. Validate Data After Extraction
```{r best_practices_validate, eval=FALSE}
# Validate extracted data
validation_checks <- list(
  record_count = nrow(demographics),
  expected_fields = all(c("record_id", "age", "gender") %in% names(demographics)),
  no_duplicate_records = nrow(demographics) == n_distinct(demographics$record_id),
  reasonable_age_range = all(demographics$age >= 0 & demographics$age <= 120, na.rm = TRUE)
)

cat("Validation Results:\n")
print(validation_checks)
```

## Troubleshooting

### Connection Issues
```{r troubleshoot_connection, eval=FALSE}
# If project creation fails, check:
# 1. URL format (should end with /api/)
# 2. Token validity
# 3. Network connectivity

tryCatch({
  project <- redcap_project_from_env()
}, error = function(e) {
  cat("Connection failed:", e$message, "\n")
  cat("Check your REDCAP_URL and REDCAP_TOKEN in .env file\n")
})
```

### Empty Results
```{r troubleshoot_empty, eval=FALSE}
# If export_records returns empty results:
result <- export_records(project, fields = c("record_id", "nonexistent_field"))

if (nrow(result) == 0) {
  cat("No data returned. Check:\n")
  cat("- Field names are correct\n")
  cat("- Records exist in the project\n")
  cat("- User has appropriate permissions\n")
  
  # Check available fields
  available_fields <- names(project$data)
  cat("Available fields:", paste(head(available_fields, 10), collapse = ", "), "...\n")
}
```

## Summary

The sardine package provides a clean, efficient way to extract REDCap data:

- **Create once, use many times**: Project objects cache data for performance
- **Fail-fast**: Connection issues are caught immediately
- **Flexible extraction**: Get all data, specific fields, records, or forms
- **Longitudinal support**: Handle complex event-based projects
- **Clean interface**: Intuitive function names and consistent API

Next, see the data import vignette to learn about getting data back into REDCap.
